A Lyapunov Analysis of Momentum Methods \\ in Optimization

### 2017-01-04 Draft14.tex
small
small changes
made several edits

\usepackage{multirow}
\usetikzlibrary{calc,arrows}
\theoremstyle{plain}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}
\newtheorem{properties}{Properties}
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
%\newtheorem{proposition}{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{observation}{Observation}
\usepackage{floatrow}
\usepackage{blindtext}
\def \endprf{\hfill {\vrule height6pt width6pt depth0pt}\medskip}
%\newenvironment{proof}{\noindent {\bf Proof} }{\endprf\par}
\usepackage[colorlinks,linkcolor=blue,citecolor=red]{hyperref}
\setlength{\tabcolsep}{18pt}
\newcommand{\vct}[1]{#1}
\newcommand{\mtx}[1]{#1}
\newcommand{\sfT}{\mathsf{T}}  % tangent space

@@

\begin{document}
\title{A Lyapunov Analysis of Momentum Methods \\ in  Optimization}
%\title{A Lyapunov Perspective on Momentum Methods \\ in  Optimization}
\date{\today}
\author{Ashia C. Wilson \and Benjamin Recht \and Michael I. Jordan}
\date{\today}
\maketitle
\begin{abstract}
Momentum methods play a central role in optimization. Several momentum methods are provably optimal, and all use a technique called {\em estimate sequences} to analyze their convergence properties.  The technique of estimate sequences has long been considered difficult to understand, leading many researchers to generate alternative, ``more intuitive'' methods and analyses. In this paper we show there is an equivalence between the technique of estimate sequences and a family of Lyapunov functions in both continuous and discrete time. This framework allows us to develop a simple and unified analysis of many existing momentum algorithms, introduce several new algorithms, and most importantly, strengthen the connection between algorithms and continuous-time dynamical systems. 
\end{abstract}

@@

To overcome the lack of intuition for the estimate sequence technique, several authors have recently proposed schemes to achieve acceleration without appealing to estimate sequences~\cite{Fazel, BubeckLeeSingh15, Lessard14}.  One promising general approach to the analysis of acceleration has been to analyze the continuous-time limit of accelerated methods and to show that the stability of the resulting ODEs can be verified via a Lyapunov function~\cite{SuBoydCandes14,Krichene15,Acceleration}.  However, these methods stop short of providing principles for deriving a discrete-time optimization algorithm from a continuous-time ODE.  There are many ways to discretize ODEs, but not all of them give rise to convergent methods or to acceleration.  Indeed, for unconstrained optimization on Euclidean spaces in the setting where the objective is strongly convex, Polyak's Heavy Ball method and Nesterov's accelerated gradient descent have the same continuous-time limit.

@@

The paper proceeds as follows.  We first describe the related work in viewing optimization algorithms as discretizations of continuous-time dynamical systems.  We then introduce the Lyapunov functions that we will use throughout the paper to demonstrate convergence of algorithms.  In particular, we will introduce time-varying Lyapunov functions that not only establish algorithm convergence but additionally verify a \emph{convergence rate}.  
%We review how such Lyapunov analysis has been used to analyze mirror decent, reviewing the concepts of Bregman divergence and mirror maps in the process. 
 We then turn to various momentum-based methods, providing a general Lyapunov analysis for Nesterov's accelerated method, the quasi-monotone subgradient method, the conditional gradient method, and a few novel algorithms.  We pay particular attention to the strongly convex case in Euclidean space, highlighting some unique properties of this setup.  Finally, we describe the connection between estimate sequences and Lyapunov functions and discuss directions for future investigation.

@@

\paragraph{Problem setting}
This paper is concerned with the class of constrained optimization problems
\begin{align}\label{eq:main-problem}
\min_{x \in \X} \; f(x),
\end{align}
where $\X \subseteq \R^d$ is a closed convex set and $f \colon \X \to \R$ is a continuously differentiable convex function. 
% We primarily focus on the case $\X = \R^d$ unless explicitly noted otherwise. 
%We also assume $f$ has a unique minimizer, $x^* \in \X$, satisfying the optimality condition $\nabla f(x^*) = 0$. 
We use the standard Euclidean norm $\|x\| = \langle x,x \rangle^{1/2}$ throughout.
We consider the general non-Euclidean setting in which the space $\X$ is endowed with a distance-generating function $h \colon \X \to \R$ that is convex and essentially smooth (i.e., $h$ is continuously differentiable in $\X$, and $\|\nabla h(x)\|_* \to \infty$ as $\|x\| \to \infty$). The function $h$ can be used to define a measure of distance in $\X$ via its Bregman divergence:
\begin{align*}
D_h(y,x) = h(y) - h(x) - \langle \nabla h(x), y-x \rangle,
\end{align*}
which is nonnegative since $h$ is convex. 
The \emph{Euclidean setting} is obtained when $h(x) = \frac{1}{2} \|x\|^2$.

@@

Krichene, Bartlett and Bayen~\cite{Krichene15} derived a modified accelerated mirror descent algorithm using a ``discretized'' Lyapunov function. However, their proof targeted a different algorithm than the algorithm in Nesterov's original work~\cite{Nesterov05}. In particular, their analysis entailed an additional smoothness assumption which is not necessary for the original algorithm~\cite{Nesterov05}, or for the original proof technique using estimate sequences.\footnote{In particular, they assume $\frac{\ell }{2} \|x-y\|^2 \leq  D_h(x,y) \leq \frac{L}{2}\|x-y\|^2$%, so that for all intensive purposes, their setting is given by the norm
. This additional assumption greatly simplifies the proof. However, only the assumption of an upper bound is needed for the most general form of accelerated methods is that $\frac{\ell }{2} \|x-y\|^2 \leq  D_h(x,y)$.} See also Nemirovski and Yudin's original work~\cite{NemirovskiiYudin} for a continuous-time perspective on mirror descent and a Lyapunov analysis of convergence.

@@

\paragraph{General accelerated mirror descent} In Wibisono {\em et al.}~\cite{Acceleration}, the following Lyapunov function:
\begin{align}\label{Eq:E}
\E_t \,=\, e^{\beta_t} (f(X_t)-f(x^*)) + D_h\left(x^*, \, X_t + e^{-\alpha_t} \dot X_t\right),
\end{align}
was introduced to analyze the flow that forms the basis for accelerated mirror descent.
%  Notice, assuming we start from rest $\dot X_0 = 0$, the  Lyapunov property allows us to conclude,
 %\begin{equation*}
 %f(X_t) - f(x^\ast)\leq \frac{e^{\beta_0} (f(X_0) - f(x^\ast) )+ D_h(x^\ast, X_0)}{e^{\beta_t}}.
 %\end{equation*}
In Section~\ref{Sec:MomLyap}, we show how this Lyapunov function emerges from a general framework for analyzing momentum dynamics; furthermore, we show how to discretize the Lyapunov function and thereby obtain a tool which can be used to analyze a variety of algorithms that are discretizations of momentum dynamics.
%MJ:  I don't view this footnote as necessary; that derivation was in supplementary information \footnote{The same derivation of the Lyapunov function was added to~\cite[App. C]{Acceleration}}
%
%
%
%ACCC GRAD STRONG
\paragraph{Accelerated gradient descent (strong convexity)}
In this paper, we show that the following function 
   \begin{equation}\label{eq:general-sc-lyap}
  \E_t =  e^{\beta_t} \left(f(X_t) - f(x^\ast) + \mu D_h\Big(x^\ast ,X_t + e^{-\alpha_t} \dot X_t\Big)\right),
 \end{equation}
is a Lyapunov function for a family of momentum dynamics when $f$ is $\mu$-strongly convex ($D_f(y,x) \geq \mu D_h(y,x)$) and $e^{\alpha_t} = \dot \beta_t$.  In Section~\ref{Sec:StrongConv}, we derive this Lyapunov function and show how it can be used to demonstrate a linear rate of convergence for the family of accelerated gradient descent dynamics.  (For simplicity, we focus on the setting in which $h = \frac{1}{2}\|\cdot\|^2$ is Euclidean.) We also show that a particular discretization of the family of dynamics provides an algorithm for which a commensurate discretization of~\eqref{eq:general-sc-lyap} is a Lyapunov function.

%MJ:  I'd omit this paragraph; it's not going to be understood by most if any readers who haven't read the other paper, and you've already gone down the route of time-varying Lyapunov.  Also, referring to Eq.~\eqref{Eq:EL} here, when it hasn't yet been introduced, is problematic.
%We briefly mention the appearance of the $e^{-\alpha_t}$ term in~\eqref{Eq:E} and~\eqref{eq:general-sc-lyap}. Given most Lyapunov functions are time-invariant, the appearance of this term in the Lyapunov function could be considered quite mysterious. In~\cite{Acceleration2}, this scaling was studied at length. In particular, it was shown that the family of Lagrangian functionals which generates~\eqref{Eq:EL} is invariant under the action of time reparameterization. By contrast, the mirror descent dynamic, which is a gradient flow, is not invariant under time-reparameterization. %Under the ideal scaling \eqref{Eq:IdeSca}, time reparameterization 
%This invariance property helps to explain why the Lyapunov function remains valid for the momentum dynamic~\eqref{Eq:E} under the ideal scaling condition~\eqref{Eq:IdeSca}.  
%in these reparameterizations of time, however, you always have a valid lyapunov function
%This invariance property might suggest why momentum family is more natural for the purposes of optimization (that is, it might help us understand why the discrete-time implementations of the family of momentum dynamics achieves systematically better rates of convergence than the family of gradient flows).

@@

\section{Momentum Methods}
\label{Sec:MomLyap}
In this section, we introduce two families of momentum dynamics, both which are generated by functionals defined on continuous-time curves. We illustrate how to derive a Lyapunov function from these dynamics. Next, we show how these dynamics can be discretized to obtain optimization algorithms. The same discrete-time mappings can be applied to the Lyapunov function, resulting in a tool which can be used to analyze the discrete-time algorithms. We provide examples of this procedure for the quasi-monotone subgradient algorithm, accelerated gradient descent, the conditional gradient algorithm, and several other novel algorithms. 
 \subsection{Lyapunov Analysis}
We study dynamics that are based on the Euler-Lagrange equation \eqref{Eq:ELBreg} in the setting where the ideal scaling~\eqref{Eq:IdeScaBet} holds with equality~$\dot \beta_t = e^{\alpha_t}$. This implies that the Bregman Lagrangian is parameterized by $\beta_t$, which determines the convergence rate:
 \begin{subequations}\label{Eq:EL}
 \begin{align}
 Z_t &= X_t + \frac{1}{\dot \beta_t} \dot X_t \label{Eq:ELZ},\\
 \frac{d}{dt}\nabla h(Z_t) &= -\dot \beta_t e^{\beta_t} \nabla f(X_t)\label{Eq:ELH}.
 \end{align}
 \end{subequations}
We begin by demonstrating how to derive the Lyapunov function~\eqref{Eq:E} for the momentum dynamics~\eqref{Eq:EL}; this derivation is similar in spirit to the Lyapunov analysis of mirror descent by Nemirovski and Yudin.  We have:
\begin{subequations}
\begin{align}
 \frac{d}{dt}D_h\left(x, Z_t\right)  &= \frac{d}{dt}\left( h(x) - h(Z_t) - \langle \nabla h(Z_t), x- Z_t\rangle\right)  \notag\\
& = - \langle \nabla h(Z_t), \dot Z_t\rangle -\left\langle \frac{d}{dt} \nabla h(Z_t),x - Z_t\right\rangle  + \langle \nabla h(Z_t), \dot Z_t\rangle \notag\\ 
 &= -\left\langle \frac{d}{dt} \nabla h\left(Z_t\right), x - Z_t\right\rangle. \, \notag%\label{eq:apply-identity}\\
 \end{align*}
 \end{subequations}
Using this identity, we obtain the following argument:
%\begin{small}
\begin{subequations}\label{Eq:LyapAnal}
\begin{align}
 \frac{d}{dt}D_h\left(x, Z_t\right)  %&= \frac{d}{dt}\left( h(x) - h(Z_t) - \langle \nabla h(Z_t), x- Z_t\rangle\right)  \notag\\
%& = - \langle \nabla h(Z_t), \dot Z_t\rangle -\left\langle \frac{d}{dt} \nabla h(Z_t),x - Z_t\right\rangle  + \langle \nabla h(Z_t), \dot Z_t\rangle \notag\\ 
 &= -\left\langle \frac{d}{dt} \nabla h\left(Z_t\right), x - Z_t\right\rangle \, \notag\\%\label{eq:apply-identity}\\
&=   \dot \beta_t e^{\beta_t}\left\langle\nabla f(X_t), x  - X_{t} - \frac{1}{\dot \beta_t} \dot X_{s} \right\rangle  \, \label{eq:apply-dynamics}   \\
& =   \dot \beta_t e^{\beta_t} \langle \nabla f(X_t), x - X_t \rangle\, - e^{ \beta_t} \langle \nabla f(X_t), \dot X_t\rangle dt\, \notag   \\
& = \dot \beta_t e^{\beta_t} \langle \nabla f(X_t), x - X_t \rangle\,  -   \frac{d}{dt}\left(e^{ \beta_t} f(X_t)\right)\,   + \dot \beta_t e^{\beta_t}  f(X_t)  \notag\\
&= \dot \beta_t e^{\beta_t} [f(X_t)+ \langle \nabla f(X_t), x - X_t \rangle] -\frac{d}{dt}\left(e^{ \beta_t} f(X_t)\right)\,  \notag  \\ %\label{Eq:Line9}\\
& \leq  \dot \beta_t e^{\beta_t}f(x) - \frac{d}{dt}\left(e^{ \beta_t} f(X_t)\right)\, \label{eq:apply-convexity}\\
& = - \frac{d}{dt} \left(e^{\beta_t}\left(f(X_t) - f(x)\right)\right) \label{Eq:cond1}.
\end{align}
\end{subequations}
%\end{small}
Here~\eqref{eq:apply-dynamics} uses the momentum dynamics%~\eqref{Eq:EL}.
~\eqref{Eq:ELH} and~\eqref{Eq:ELZ}. 
The inequality~\eqref{eq:apply-convexity} follows from the convexity of $f$. Rearranging terms and taking $x = x^\ast$, we have shown that the function~\eqref{Eq:E} has nonpositive derivative for all $t$ and is hence a Lyapunov function for the family of momentum dynamics~\eqref{Eq:EL}.

@@

   \subsection{Momentum-Based Algorithms}

@@

E_{k+1} - E_k \leq \varepsilon_k,

@@

\varepsilon_k = -D_h(z_{k+1}, z_k).

@@

%we follow a similar argument to that in~\eqref{Eq:LyapAnal}. %Thus, following argument holds for~\eqref{Eq:AlgoForward},
%First, we define our discrete-time energy function~\eqref{Eq:E},
%\begin{align}\label{Eq:DiscLyapFunc1}
% E_k = A_k(f(x_k) - f(x)) + D_h(x, z_{k}),
% \end{align}
%which is the result of making the same continuous to discrete time identifications. Note that,

@@

%{\color{red} \frac{d}{dt}D_h(x, X_{t} + e^{-\alpha_{t}} \dot X_{t})}\,  
%& =  \int_{t_0}^{t_1} \frac{d}{ds}D_h(x, X_{s} + e^{-\alpha_{s}} \dot X_{s})\, ds +D_h(x, X_{t_0} + e^{-\alpha_{t_0}} \dot X_{t_0})\\
%& {\color{red} =  - \left\langle \frac{d}{dt} \nabla h(X_t+ e^{-\alpha_t}\dot X_t), x - X_t - e^{-\alpha_t} \dot X_t\right\rangle} \,  \notag\\

@@

%&{\color{red} \overset{\eqref{Eq:ELBreg}}{=}e^{\alpha_t + \beta_t} \langle \nabla f(X_t), x - X_t \rangle\, - e^{ \beta_t} \langle \nabla f(X_t), \dot X\rangle}\, \notag   \\

@@

%& {\color{red}= - \frac{d}{dt}\left\{ e^{\beta_t}\left(f(X_t) - f(x)\right)\right\}\label{Eq:cond1}}\\
%& = - D_h(z_{k+1}, z_k), \notag

@@

f(x_k) - f(x^\ast) \leq \frac{D_h(x^\ast, x_0) + A_0 (f(x_0) - f(x^\ast))}{A_k}.

@@

%MJ This sentence isn't clear, and I'm not sure it's worth including.
%Nevertheless, the proof structure which we demonstrated from this thought experiment serves as the backbone for analyzing many discretizations of \eqref{Eq:DynaBeta}, several of which constitute popular methods used throughout optimization.

\subsection{The Quasi-Monotone Subgradient Method}
\label{Sec:QuasiMono}
The quasi-monotone subgradient method was developed by Nesterov~\cite{Nesterov15} as an alternative to the dual averaging method.  It also achieves the $O(1/\sqrt{k})$ lower bound for the class of methods with bounded subgradients. The advantage of the quasi-monotone method however is that a convergence-rate guarantee can be shown for the entire sequence of iterates instead of the average (or minimum) iterate, as is the case with the dual averaging method in the non-Euclidean setting. {\em We begin by studying the quasi-monotone method in the situation where we have full gradients, which we assume to be bounded} and show later how the analysis extends to the setting where we only have access to bounded subgradients. We also set the dual averaging term $\gamma_k \equiv 1$ for convenience and give an analysis of the full algorithm in Appendix~\ref{App:ProofQuasiGen}.  With these minor modifications, the algorithm can be understood as the explicit Euler method applied to~\eqref{Eq:Z} and the implicit Euler method applied to~\eqref{Eq:X} (where now we take $e^{\beta_t} = A_{k+1}$, while leaving our identification for $\frac{d}{dt} e^{\beta_t}$ unchanged):

>>>>

\usepackage{multirow}
\usetikzlibrary{calc,arrows}
\theoremstyle{plain}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}
\newtheorem{properties}{Properties}
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
%\newtheorem{proposition}{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{observation}{Observation}
\usepackage{blindtext}
\def \endprf{\hfill {\vrule height6pt width6pt depth0pt}\medskip}
%\newenvironment{proof}{\noindent {\bf Proof} }{\endprf\par}
\usepackage[colorlinks,linkcolor=blue,citecolor=red]{hyperref}
\setlength{\tabcolsep}{18pt}
\newcommand{\vct}[1]{#1}
\newcommand{\mtx}[1]{#1}
\newcommand{\sfT}{\mathsf{T}}  % tangent space

@@

\begin{document}
\title{A Lyapunov Analysis of Momentum Methods \\ in  Optimization}
%\title{A Lyapunov Perspective on Momentum Methods \\ in  Optimization}
\date{\today}
\author{Ashia C. Wilson \qquad Benjamin Recht \qquad Michael I. Jordan\\ \\
University of California, Berkeley
}
\date{\today}
\maketitle
\begin{abstract}
Momentum methods play a central role in optimization. Several momentum methods are provably optimal, and all use a technique called {\em estimate sequences} to analyze their convergence properties.  The technique of estimate sequences has long been considered difficult to understand, leading many researchers to generate alternative, ``more intuitive'' methods and analyses. In this paper we show there is an equivalence between the technique of estimate sequences and a family of Lyapunov functions in both continuous and discrete time. This framework allows us to develop a simple and unified analysis of many existing momentum algorithms, introduce several new algorithms, and most importantly, strengthen the connection between algorithms and continuous-time dynamical systems. 
\end{abstract}

@@

To overcome the lack of intuition for the estimate sequence technique, several authors have recently proposed schemes to achieve acceleration without appealing to estimate sequences~\cite{Fazel, BubeckLeeSingh15, Lessard14, DroriTeboulle13}.  One promising general approach to the analysis of acceleration has been to analyze the continuous-time limit of accelerated methods and to show that the stability of the resulting ODEs can be verified via a Lyapunov function~\cite{SuBoydCandes14,Krichene15,Acceleration}.  However, these methods stop short of providing principles for deriving a discrete-time optimization algorithm from a continuous-time ODE.  There are many ways to discretize ODEs, but not all of them give rise to convergent methods or to acceleration.  Indeed, for unconstrained optimization on Euclidean spaces in the setting where the objective is strongly convex, Polyak's Heavy Ball method and Nesterov's accelerated gradient descent have the same continuous-time limit.

@@

The paper proceeds as follows.  We first describe the related work in viewing optimization algorithms as discretizations of continuous-time dynamical systems.  We then introduce the Lyapunov functions that we will use throughout the paper to demonstrate convergence of algorithms.  In particular, we will introduce time-varying Lyapunov functions that not only establish algorithm convergence but additionally verify a \emph{convergence rate}.  
 We then turn to various momentum-based methods, providing a general Lyapunov analysis for Nesterov's accelerated method, the quasi-monotone subgradient method, the conditional gradient method, and a few novel algorithms.  We pay particular attention to the strongly convex case in Euclidean space, highlighting some unique properties of this setup.  Finally, we describe the connection between estimate sequences and Lyapunov functions and discuss directions for future investigation.

@@

\paragraph{Problem setting}
This paper is concerned with the class of constrained optimization problems
\begin{align}\label{eq:main-problem}
\min_{x \in \X} \; f(x),
\end{align}
where $\X \subseteq \R^d$ is a closed convex set and $f \colon \X \to \R$ is a continuously differentiable convex function. 
We use the standard Euclidean norm $\|x\| = \langle x,x \rangle^{1/2}$ throughout.
We consider the general non-Euclidean setting in which the space $\X$ is endowed with a distance-generating function $h \colon \X \to \R$ that is convex and essentially smooth (i.e., $h$ is continuously differentiable in $\X$, and $\|\nabla h(x)\|_* \to \infty$ as $\|x\| \to \infty$). The function $h$ can be used to define a measure of distance in $\X$ via its Bregman divergence:
\begin{align*}
D_h(y,x) = h(y) - h(x) - \langle \nabla h(x), y-x \rangle,
\end{align*}
which is nonnegative since $h$ is convex. 
The \emph{Euclidean setting} is obtained when $h(x) = \frac{1}{2} \|x\|^2$.

@@

Krichene, Bartlett and Bayen~\cite{Krichene15} derived a modified accelerated mirror descent algorithm using a ``discretized'' Lyapunov function. However, their proof targeted a different algorithm than the algorithm in Nesterov's original work~\cite{Nesterov05}. In particular, their analysis entailed an additional smoothness assumption which is not necessary for the original algorithm~\cite{Nesterov05}, or for the original proof technique using estimate sequences.\footnote{In particular, they assume $\frac{\ell }{2} \|x-y\|^2 \leq  D_h(x,y) \leq \frac{L}{2}\|x-y\|^2$%, so that for all intensive purposes, their setting is given by the norm
. This additional assumption greatly simplifies the proof. However, only the assumption of an upper bound, $\frac{\ell }{2} \|x-y\|^2 \leq  D_h(x,y)$,  is needed for the most general form of accelerated methods.} See also Nemirovski and Yudin's original work~\cite{NemirovskiiYudin} for a continuous-time perspective on mirror descent and a Lyapunov analysis of convergence.

@@

\paragraph{General accelerated mirror descent} In Wibisono {\em et al.}~\cite{Acceleration}, the following Lyapunov function:
\begin{align}\label{Eq:E}
\E_t \,=\, e^{\beta_t} (f(X_t)-f(x^*)) + D_h\left(x^*, \, X_t + e^{-\alpha_t} \dot X_t\right),
\end{align}
was introduced to analyze the flow that forms the basis for accelerated mirror descent.
In Section~\ref{Sec:MomLyap}, we show how this Lyapunov function emerges from a general framework for analyzing momentum dynamics; furthermore, we show how to discretize the Lyapunov function and thereby obtain a tool which can be used to analyze a variety of algorithms that are discretizations of momentum dynamics.

\paragraph{Accelerated gradient descent (strong convexity)}
In this paper, we show that the following function 
   \begin{equation}\label{eq:general-sc-lyap}
  \E_t =  e^{\beta_t} \left(f(X_t) - f(x^\ast) + \mu D_h\Big(x^\ast ,X_t + e^{-\alpha_t} \dot X_t\Big)\right),
 \end{equation}
is a Lyapunov function for a family of momentum dynamics when $f$ is $\mu$-strongly convex ($D_f(y,x) \geq \mu D_h(y,x)$).  In Section~\ref{Sec:StrongConv}, we derive this Lyapunov function and show how it can be used to demonstrate a linear rate of convergence for the family of accelerated gradient descent dynamics.  (For simplicity, we focus on the setting in which $h = \frac{1}{2}\|\cdot\|^2$ is Euclidean.) We also show that a particular discretization of the family of dynamics provides an algorithm for which a commensurate discretization of~\eqref{eq:general-sc-lyap} is a Lyapunov function.

@@

\section{Momentum Methods}
\label{Sec:MomLyap}
In this section, we introduce two families of momentum dynamics, both of which are generated by functionals defined on continuous-time curves. We illustrate how to derive a Lyapunov function from these dynamics. Next, we show how these dynamics can be discretized to obtain optimization algorithms. The same discrete-time mappings can be applied to the Lyapunov function, resulting in a tool which can be used to analyze the discrete-time algorithms. We provide examples of this procedure for the quasi-monotone subgradient algorithm, accelerated gradient descent, the conditional gradient algorithm, and several other novel algorithms. 
 \subsection{Lyapunov analysis}
We study dynamics that are based on the Euler-Lagrange equation \eqref{Eq:ELBreg} in the setting where the ideal scaling~\eqref{Eq:IdeScaBet} holds with equality~$\dot \beta_t = e^{\alpha_t}$. This implies that the Bregman Lagrangian is parameterized by $\beta_t$, which determines the convergence rate:
 \begin{subequations}\label{Eq:EL}
 \begin{align}
 Z_t &= X_t + \frac{1}{\dot \beta_t} \dot X_t \label{Eq:ELZ},\\
 \frac{d}{dt}\nabla h(Z_t) &= -\dot \beta_t e^{\beta_t} \nabla f(X_t)\label{Eq:ELH}.
 \end{align}
 \end{subequations}
We begin by demonstrating how to derive the Lyapunov function~\eqref{Eq:E} for the momentum dynamics~\eqref{Eq:EL}; this derivation is similar in spirit to the Lyapunov analysis of mirror descent by Nemirovski and Yudin.  We have:
\begin{subequations}
\begin{align*}
 \frac{d}{dt}D_h\left(x, Z_t\right)  &= \frac{d}{dt}\left( h(x) - h(Z_t) - \langle \nabla h(Z_t), x- Z_t\rangle\right)  \notag\\
& = - \langle \nabla h(Z_t), \dot Z_t\rangle -\left\langle \frac{d}{dt} \nabla h(Z_t),x - Z_t\right\rangle  + \langle \nabla h(Z_t), \dot Z_t\rangle \notag\\ 
 &= -\left\langle \frac{d}{dt} \nabla h\left(Z_t\right), x - Z_t\right\rangle. \, \notag%\label{eq:apply-identity}\\
 \end{align*}
 \end{subequations}
Using this identity, we obtain the following argument:
%\begin{small}
\begin{subequations}\label{Eq:LyapAnal}
\begin{align}
 \frac{d}{dt}D_h\left(x, Z_t\right)  
 &= -\left\langle \frac{d}{dt} \nabla h\left(Z_t\right), x - Z_t\right\rangle \, \notag\\%\label{eq:apply-identity}\\
&=   \dot \beta_t e^{\beta_t}\left\langle\nabla f(X_t), x  - X_{t} - \frac{1}{\dot \beta_t} \dot X_{s} \right\rangle  \, \label{eq:apply-dynamics}   \\
& =   \dot \beta_t e^{\beta_t} \langle \nabla f(X_t), x - X_t \rangle\, - e^{ \beta_t} \langle \nabla f(X_t), \dot X_t\rangle dt\, \notag   \\
& = \dot \beta_t e^{\beta_t} \langle \nabla f(X_t), x - X_t \rangle\,  -   \frac{d}{dt}\left(e^{ \beta_t} f(X_t)\right)\,   + \dot \beta_t e^{\beta_t}  f(X_t)  \notag\\
&= \dot \beta_t e^{\beta_t} [f(X_t)+ \langle \nabla f(X_t), x - X_t \rangle] -\frac{d}{dt}\left(e^{ \beta_t} f(X_t)\right)\,  \notag  \\ %\label{Eq:Line9}\\
& \leq  \dot \beta_t e^{\beta_t}f(x) - \frac{d}{dt}\left(e^{ \beta_t} f(X_t)\right)\, \label{eq:apply-convexity}\\
& = - \frac{d}{dt} \left(e^{\beta_t}\left(f(X_t) - f(x)\right)\right) \label{Eq:cond1}.
\end{align}
\end{subequations}
%\end{small}
Here~\eqref{eq:apply-dynamics} uses the momentum dynamics%~\eqref{Eq:EL}.
~\eqref{Eq:ELH} and~\eqref{Eq:ELZ}. 
The inequality~\eqref{eq:apply-convexity} follows from the convexity of $f$. Rearranging terms and taking $x = x^\ast$, we have shown that the function~\eqref{Eq:E} has nonpositive derivative for all $t$ and is hence a Lyapunov function for the family of momentum dynamics~\eqref{Eq:EL}.

@@

   \subsection{Momentum-based algorithms}

@@

E_{k+1} - E_k \leq \varepsilon_{k+1},

@@

\varepsilon_{k+1} = -D_h(z_{k+1}, z_k).

@@

@@

@@

@@

@@

f(x_k) - f(x^\ast) \leq \frac{ A_0 (f(x_0) - f(x^\ast)) + D_h(x^\ast, x_0) }{A_k}.

@@

\subsection{The Quasi-monotone subgradient method}
\label{Sec:QuasiMono}
The quasi-monotone subgradient method was developed by Nesterov~\cite{Nesterov15} as an alternative to the dual averaging method.  It also achieves the $O(1/\sqrt{k})$ lower bound for the class of methods with bounded subgradients. The advantage of the quasi-monotone method however is that a convergence-rate guarantee can be shown for the entire sequence of iterates instead of the average (or minimum) iterate, as is the case with the dual averaging method in the non-Euclidean setting. {\em We begin by studying the quasi-monotone method in the situation where we have full gradients, which we assume to be bounded} and show later how the analysis extends to the setting where we only have access to bounded subgradients. We also set the dual averaging term $\gamma_k \equiv 1$ for convenience and give an analysis of the full algorithm in Appendix~\ref{App:ProofQuasiGen}.  With these minor modifications, the algorithm can be understood as the explicit Euler method applied to~\eqref{Eq:Z} and the implicit Euler method applied to~\eqref{Eq:X} (where now we take $e^{\beta_t} = A_{k+1}$, while leaving our identification for $\frac{d}{dt} e^{\beta_t}$ unchanged):

\begin{algorithm}[H]
\caption{The Quasi-Monotone (Sub)gradient Method $\gamma_k \equiv 1$}
{\bf Assumptions:} $f, h$ are convex and differentiable. $h$ is strongly convex and $f$ has bounded gradients, $\sup_{x\in\X}\|\nabla f(x)\| = G <\infty$. \\
Let $A_0 = 1$, $x_0 = z_0$, $\tau_k = \frac{A_{k+1} - A_k}{A_{k+1}}$, $\alpha_k = A_{k} - A_{k-1}$ . Define recursively,
\begin{subequations}\label{Eq:QuasiSub1}
\begin{align}
x_{k+1} &= \tau_k z_k + (1 - \tau_k)x_{k} ,\label{Eq:ZSeqQuasi1}\\
z_k &= \arg \min_{z\in \X} \left\{ \sum_{i=1}^k \alpha_i \langle \nabla f(x_i), z\rangle + D_h(z, x_0)\right\} \label{Eq:XSeqQuasi1}.
\end{align} 
\end{subequations}
\end{algorithm}
\noindent This modified quasi-monotone subgradient method has optimality conditions
\begin{subequations}\label{Eq:QuasiSub}
\begin{align}
z_{k} &= x_{k} + \frac{A_{k+1}}{A_{k+1} -A_{k}} (x_{k+1} - x_k),\label{Eq:ZSeqQuasi}\\
\nabla h(z_{k+1}) -\nabla h(z_k) &= -(A_{k+1} - A_{k})\nabla f(x_{k+1})\label{Eq:XSeqQuasi}.
\end{align} 
\end{subequations}
%%%%%%%%%%%%%%%%%%%%%%% Estimate Sequences

The following result illustrates how the discrete Lyapunov function~\eqref{Eq:DiscLyapFunc1} can be used to analyze~\eqref{Eq:QuasiSub}:
\begin{theorem}
\label{Thm:Quasi}
Using the Lyapunov function~\eqref{Eq:DiscLyapFunc1},
we can show the quasi-monotone method~\eqref{Eq:QuasiSub} satisfies \[E_{k+1} - E_k \leq \,\,\varepsilon_{k+1},\] where the error scales in the following way,%\footnote{For reference, the error for subgradient descent scales in the same way (see Appendix~\ref{App:MirrorDesc})}
\[\varepsilon_{k+1} = \frac{(A_{k+1}- A_k)^2}{2}\|\nabla f(x_{k+1})\|^2 \leq \frac{(A_{k+1}- A_k)^2}{2}G^2.\]
\end{theorem}
\noindent Taking $x = x^\ast$ and summing over $k$ results in the following convergence rate:
\begin{align}\label{Eq:BoundQuasi}
f(x_k) - f(x^\ast) \leq \frac{A_0 (f(x_0) - f(x^\ast))+D_h(x^\ast , x_0)  + \frac{1}{2}\sum_{i=1}^{k}\varepsilon_i }{A_k}.
\end{align}
If we optimize the bound~\eqref{Eq:BoundQuasi} over $A_k$, we can obtain an $O(1/\sqrt{k})$ convergence rate guarantee by setting a time-horizon for the algorithm to be run, and choosing $A_{k+1} - A_{k} =\frac{D_h(x,x_0)}{G\sqrt{k+1}}.$ Without this step, we suffer an additional $\log k$ factor in the the numerator.\footnote{See~\cite[2]{Nesterov15} for more details on this history on subgradient methods as well as step-size analysis.}
